---
title: "Reds Project"
author: "Jake Federman"
date: "2023-10-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(tidyverse)
```

```{r}
df <- read_csv(file = "data.csv")
```

# Hypothesis

Dew point affects pitches in the following ways:

**Enhances downward trajectory on topspin pitches**

As a result, a pitch with topspin at a high dew point might come in at a steeper approach angle than an identical pitch thrown at a low dew point. We can build a model to predict vertical approach angle from the factors that contribute to it, and attribute unexplained variance in the steeper direction to dew point.

**Enhances horizontal movement on sidespin pitches**

As a result, a pitch with sidespin at a high dew point might come in at a more extreme horizontal approach angle than an identical pitch thrown at a low dew point. We can build a model to predict horizontal approach angle from the factors that contribute to it, and attribute unexplained variance in the extreme direction to dew point.


# Model for Vertical Approach Angle on topspin pitches:

Vertical approach angle is a product of:

* **Velocity:** The faster the pitch comes in, the less time gravity has to bring it down
* **Vertical Release Height:** Pitches released higher will be steeper, and pitches released lower will be flatter
* **Induced Vertical Break:** Pitches with more carry will come in at a flatter angle
* **Pitch Location:** The higher the pitch is, the flatter it will be. The lower it is, the steeper it will be

```{r}
#Filter the dataset, only include topspin pitches (induced_vertical_break < 0)
df_topspin <- df %>% filter(INDUCED_VERTICAL_BREAK < 0)
#Build the model
model_vaa_topspin <- lm(data = df_topspin, VERTICAL_APPROACH_ANGLE ~ 
                     RELEASE_SPEED + INDUCED_VERTICAL_BREAK + 
                     RELEASE_HEIGHT + PLATE_Z)

predicted_vaa <- predict(model_vaa_topspin, newdata = df_topspin)

# Calculate residuals
residuals <- df_topspin$VERTICAL_APPROACH_ANGLE - predicted_vaa

# Compute the standard deviation of the residuals
residual_sd <- sd(residuals)

# Compute z-scores for each observation
z_scores <- residuals / residual_sd

# Compute the one-tailed probability for each z-score in the direction of more negative
probabilities <- pnorm(z_scores, lower.tail = TRUE)  # TRUE gives the left tail

# Create a data frame to store the results
results <- data.frame(
  PID = df_topspin$PID,
  predicted_vaa = predicted_vaa,
  actual_vaa = df_topspin$VERTICAL_APPROACH_ANGLE,
  prob_affected_dew_point_vert = 1 - probabilities
)

# View the results
head(results)
```

# Model for horizontal approach angle

Horizontal approach angle is a product of:

* **Release Side:** Pitches released wider will approach from wider
* **Horizontal Break:** Pitches with more horizontal movement will come in at a more extreme angle
* **Pitch Location:** The further the pitch location from the release side, the more extreme the approach angle will be

```{r}
#Build the model
model_haa_sidespin <- lm(data = df, HORIZONTAL_APPROACH_ANGLE ~ 
                           RELEASE_SIDE + HORIZONTAL_BREAK + PLATE_X)

predicted_haa <- predict(model_haa_sidespin, newdata = df)

# Calculate residuals
residuals_2 <- df$HORIZONTAL_APPROACH_ANGLE - predicted_haa

# Compute the standard deviation of the residuals
residual_sd_2 <- sd(residuals_2)

# Compute z-scores for each observation
z_scores_2 <- residuals_2 / residual_sd_2

tail <- ifelse(df$HORIZONTAL_BREAK < 0, TRUE, FALSE)  
  # TRUE for left tail, FALSE for right tail

# Compute the one-tailed probability for each z-score based on the value of HORIZONTAL_BREAK
probabilities_2 <- mapply(function(z, tail) pnorm(z, lower.tail = tail), z_scores_2, tail)

# Create a data frame to store the results
results_2 <- data.frame(
  PID = df$PID,
  predicted_haa = predicted_haa,
  actual_haa = df$HORIZONTAL_APPROACH_ANGLE,
  prob_affected_dew_point_horiz = 1 - probabilities_2
)

# View the results
head(results_2)
```
# Final Analysis

```{r}
df <-
  df %>% 
  left_join(results_2, by = "PID") %>% 
  select(PID, prob_affected_dew_point_horiz) %>% 
  left_join(results, by = "PID") %>% 
  mutate(prob_affected_dew_point_vert =
           ifelse(is.na(prob_affected_dew_point_vert), 0.5, #Assign probability
                                                            #0.5 to all pitches w/ IVB > 0
                  prob_affected_dew_point_vert)) %>% 
  select(PID, prob_affected_dew_point_vert, prob_affected_dew_point_horiz)

```
```{r}
#P(pitch was affected) = = 1 - (P(not affected vertically)*P(not affected horizontally))
df <-
  df %>% 
  mutate(DEWPOINT_AFFECTED = 
           1-((1-prob_affected_dew_point_vert)*(1-prob_affected_dew_point_horiz))) %>% 
  select(PID, DEWPOINT_AFFECTED)
head(df)
```

```{r}
write_csv(df, file = "submission.csv")
```